By solving the following exercises, you can practice the material discussed in this chapter. Solutions to all exercises are available with the code download on the book’s website at \url{www.wiley.com/go/proc++6e}. However, if you are stuck on an exercise, first reread parts of this chapter to try to find an answer yourself before looking at the solution from the website.

\begin{itemize}
\item
Exercise 22-1: Let’s play a bit with durations. Create a duration, d1, with a precision of seconds, initialized to 42 seconds. Create a second duration, d2, with a precision of minutes, initialized to 1.5 minutes. Calculate the sum of d1 and d2. Write out the result to the standard output, once expressed in seconds, once expressed in minutes.

\item
Exercise 22-2: Ask the user to enter a date as yyyy-mm-dd, for example, 2020-06-22. Use a regular expression (see Chapter 21) to extract the year, month, and day components, and finally, use year\_month\_day to validate the date.

\item
Exercise 22-3: Write a getNumberOfDaysBetweenDates() function that calculates the number of days between two given dates. Test your implementation in your main() function.

\item
Exercise 22-4: Write a program that prints out the day of the week of June 22, 2020.

\item
Exercise 22-5: Construct a UTC time. Convert this time to the local time in Tokyo, Japan. Further convert the resulting time to New York time. And finally convert the resulting time to GMT. Verify that the original UTC time and the final GMT time are equal. Tip: The time zone identifier for Tokyo is Asia/Tokyo, for New York it is America/New\_York, and for GMT it is GMT.

\item
Exercise 22-6: Write a getDurationSinceMidnight() function that returns the duration between midnight and the current local time in seconds. Use your function to print out the number of seconds since midnight to the standard output console. Finally, use the hh\_mm\_ss class to convert the duration returned by your function to hours, minutes, and seconds, and print the result on standard output.
\end{itemize}











