\begin{itemize}
\item
Robert C. Martin. Agile Software Development, Principles, Patterns, and Practices. Pearson, 2003. ISBN: 978-1292025940.

Written for software engineers “in the trenches,” this text focuses on the technology—the principles, patterns, and process—that help software engineers effectively manage increasingly complex operating systems and applications.

\item
Mike Cohn. Succeeding with Agile: Software Development Using Scrum. Addison-Wesley Professional, 2009. ISBN: 0-321-57936-4.

An excellent guide to start with the Scrum methodology.

\item
David Thomas and Andrew Hunt, The Pragmatic Programmer, Your Journey to Mastery, 2nd ed. Addison-Wesley Professional, 2019. ISBN: 978-0135957059.

A new edition of a classic book, and a must-read for every software engineer. More than 20 years later, the advice of the first edition is still spot on. It examines the core process—what do you do, as an individual and as a team, if you want to create software that’s easy to work with and good for your users.

\item
Barry W. Boehm, TRW Defense Systems Group. A Spiral Model of Software Development and Enhancement. IEEE Computer, 21(5): 61–72, 1988.

A landmark paper that described the state of software development at the time and proposed the spiral model.

\item
Kent Beck and Cynthia Andres. Extreme Programming Explained: Embrace Change, 2nd ed.

Addison-Wesley Professional, 2004. ISBN: 0-321-27865-8.
One of several books in a series that promote Extreme Programming as a new approach to software development.

\item
Robert T. Futrell, Donald F. Shafer, and Linda Isabell Shafer. Quality Software Project Management. Prentice Hall, 2002. ISBN: 0-130-91297-2.

A guidebook for anybody who is responsible for the management of software development processes.

\item
Robert L. Glass. Facts and Fallacies of Software Engineering. Addison-Wesley Professional, 2002. ISBN: 0-321-11742-5.

Discusses various aspects of the software development process and exposes hidden truisms along the way.

\item
Philippe Kruchten. The Rational Unified Process: An Introduction, 3rd ed. Addison-Wesley Professional, 2003. ISBN: 0-321-19770-4.

Provides an overview of RUP, including its mission and processes.

\item
Edward Yourdon. Death March, 2nd ed. Prentice Hall, 2003. ISBN: 0-131-43635-X.

A wonderfully enlightening book about the politics and realities of software development.

\item
Wikipedia contributors. Scrum. Wikipedia, The Free Encyclopedia, \url{en.wikipedia.org/wiki/Scrum_(software_development)} (accessed August 15, 2023).

A detailed discussion of the Scrum methodology.

\item
Manifesto for Agile Software Development. \url{agilemanifesto.org} (accessed August 15, 2023).

The complete agile software development manifesto.

\item
Wikipedia contributors. Version control. Wikipedia, The Free Encyclopedia, \url{en.wikipedia.org/wiki/Version_control} (accessed August 15, 2023).

Explains the concepts behind revision control systems and what kinds of solutions are available.
\end{itemize}