
\begin{itemize}
\item
Bjarne Stroustrup and Herb Sutter. C++ Core Guidelines. \url{github.com/isocpp/CppCoreGuidelines/blob/master/CppCoreGuidelines.md} (accessed August 15, 2023).

This document is a set of guidelines for using C++ well. The aim of this document is to help people to use modern C++ effectively.

\item
Martin Fowler. Refactoring: Improving the Design of Existing Code, 2nd ed. Addison-Wesley Professional, 2018. ISBN: 0-134-75759-9.

A new edition of a classic book that espouses the practice of recognizing and improving bad code.

\item
Herb Sutter and Andrei Alexandrescu. C++ Coding Standards: 101 Rules, Guidelines, and Best Practices. Addison-Wesley Professional, 2004. ISBN: 0-321-11358-0.

A must-have book on C++ design and coding style. “Coding standards” here doesn’t mean “how many spaces I should indent my code.” This book contains 101 best practices, idioms, and common pitfalls that can help you to write correct, understandable, and efficient C++ code.

\item
Diomidis Spinellis. Code Reading: The Open Source Perspective. Addison-Wesley Professional, 2003. ISBN: 0-201-79940-5.

A unique book that turns the issue of programming style upside down by challenging the reader to learn to read code properly in order to become a better programmer.

\item
Dimitri van Heesch. Doxygen. www.doxygen.nl (accessed August 15, 2023).

A highly configurable program that generates documentation from source code and comments.

\item
John Aycock. Reading and Modifying Code. John Aycock, 2008. ISBN 0-980-95550-5.

A nice little book with advice about how to perform the most common operations on code: reading, modifying, testing, debugging, and writing.

\item
Wikipedia contributors. Code refactoring. Wikipedia, The Free Encyclopedia, \url{en.wikipedia.org/wiki/Refactoring} (accessed August 15, 2023).

A discussion of what code refactoring means, including a number of techniques for refactoring.

\item
Google. Google C++ Style Guide. \url{google.github.io/styleguide/cppguide.html} (accessed August 15, 2023).

A discussion of the C++ style guidelines used at Google.
\end{itemize}


