\noindent
\textbf{WHAT’S IN THIS CHAPTER?}

\begin{itemize}
\item
How to write class templates

\item
How the compiler processes templates

\item
How to organize template source code

\item
How to use non-type template parameters

\item
How to write templates of individual class member functions

\item
How to customize class templates for specific types

\item
How to combine templates and inheritance

\item
How to write alias templates

\item
How to write function templates

\item
How to make function templates friends of class templates

\item
How to use the abbreviated function template syntax

\item
How to use variable templates

\item
What concepts are

\item
How to use concepts to type constrain auto

\item
How to use concepts to specify requirements for template type parameters
\end{itemize}

\noindent
\textbf{WILEY.COM DOWNLOADS FOR THIS CHAPTER}

Please note that all the code examples for this chapter are available as part of this chapter’s code download on the book’s website at \url{www.wiley.com/go/proc++6e} on the Download Code tab.

C++ provides language support not only for object-oriented programming but also for generic programming. As discussed in Chapter 6, “Designing for Reuse,” the goal of generic programming is to write reusable code. The fundamental tools for generic programming in C++ are templates. Although not strictly an object-oriented feature, templates can be combined with object-oriented programming for powerful results. Using existing templates, such as those provided by the Standard Library, e.g., std::vector, unique\_ptr, and so on, is usually straightforward. However, many programmers consider writing their own templates to be the most difficult part of C++ and, for that reason, tend to avoid writing them. However, as a professional C++ programmer, you need to know how to write class and function templates.

This chapter provides the coding details for fulfilling the design principle of generality discussed in Chapter 6, while Chapter 26, “Advanced Templates,” delves into some of the more advanced template features.


















